---
header-includes: \usepackage{color}
                  \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## General Info for Homeworks

-  if you don't follow the instructions given below it <font color="red)WILL</font> cost you points!

-  any library that you use in your Rmd has to be loaded at the beginning.

-  call the file *hw# .Rmd*, where #  is the number, for example *hw1.Rmd* (but NOT homework1.Rmd, Hw1.Rmd, HW1.Rmd, hw1.rmd etc)

-  put it in your shared folder, not in any subfolder

-  make sure it knits on <font color="red) **MY** </font> computer.

-  if you read in anything (Resma3.RData, a text file, anything) make sure to read it from the class webpage, not from you hard drive. For example, have

```{r}
load(url("http://academic.uprm.edu/wrolke/Resma3/Resma3.RData"))
```

and not 

```{r eval=FALSE}
load("C:/myfolder/Resma3.RData"))
```

it is ok to have that while you are working on the assignment, but you have to change it back before you submit it.

-  if you have a chunk that won't knit but yields an error, but you want to have it as part of your solution use the following:

\`\`\`{r error=TRUE}

-  Starting with Ver 1.2 RStudio needs a space between a hashtag(s) and the text for a header (this is so a hashtag can be used as a hashtag)

-  unless I say differently your routines should always be able to handle a general case, not just one specific example. If the assignment includes an example, it might be a good idea if you make one up yourself and write your routine for that. Then when you think you got it apply your routine to the example from the assignment.

so for example say the problem statement is: draw the following graph:

```{r}
attach(wine)
plot(Heart.Disease.Deaths, Wine.Consumption)
```

your own example might be:

```{r}
Amount <- runif(20, 1, 6)
Dead <- 100*Amount + runif(20, 0, 100)
plot(Amount, Dead)
```

-  do not use any libraries that have not been discussed in class. If you want to do so send me an email and ask first.
