---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## R Basics I

To start run

```{r, eval=FALSE}
ls() 
```

This shows you a "listing"" of the files (data, routines etc.) in the current project. (Likely there is nothing there right now)

Everything in R is either a data set or a function. It is a function if it is supposed to do something (maybe calculate something, show you something like a graph or something else etc. ). If it is a function is ALWAYS NEEDS (). Sometimes the is something in between the parentheses, like in 
```{r echo=FALSE}
x <- 1:11
```

```{r}
mean(x)
```

Sometimes there isn't like in the ls(). But the () has to be there anyway.

If you have worked for a while you might have things you need to save, do that by clicking on 

File > Save 

RStudio has a nice recall feature, using the up and down arrow keys. Also, clicking on the History tab shows you the recently run commands. Finally, typing the first three letters of a command in the console and then typing CTRL-^  shows you a list of when you ran commands like this the last times.

R is case-sensitive, so a and A are two different things.

Often during a session you create objects that you need only for a short time. When you no longer need them use  **rm** to get rid of them:
  
```{r}
x <- 10
x^2
rm(x) 
```

the **<-** is the *assignment* character in R, it assigns what is on the right to the symbol on the left. (Think of an arrow to the left)

###  Data Entry 

For a few numbers the easiest thing is to just type them in:
```{r}
x <-  c(10, 2, 6, 9)
x
```
c() is a function that takes the objects inside the () and  **c**ombines them into one single object (a vector). 

### Data Types in R

the most basic type of data in R is a **vector**, simply a list of values. 

Say we want the numbers 1.5, 3.6, 5.1 and 4.0 in an R vector called x, then we can type 

```{r}
x <- c(1.5, 3.6, 5.1, 4.0)
x
```

Often the numbers have a structure one can make use of:

``` {r}
1:10 
10:1
1:20*2
c(1:10, 1:10*2)
```

Sometimes you need parentheses:

```{r}
n <- 10
1:n-1
1:(n-1)
```

The *rep* ("repeat") command is very useful:

```{r}
rep(1, 10)
rep(1:3, 10)
rep(1:3, each=3)
rep(c("A", "B", "C"), c(4,7,3))
```

what does this do?
```{r, eval=FALSE}
rep(1:10, 1:10)
```

### Commands for Vectors

To find out how many elements a vector has use the *length* command: 

```{r}
x <- c(1.4, 5.1, 2.0, 6.8, 3.5, 2.1, 5.6, 3.3, 6.9, 1.1)
length(x)
```

The elements of a vector are accessed with the bracket [ ] notation:

```{r}
x[3]
x[1:3]
x[c(1, 3, 8)]
x[-3]
x[-c(1, 2, 5)]
```

Instead of numbers a vector can also consist of characters (letters, numbers, symbols etc.) These are identified by quotes:

````{r} 
c("A", "B", 7, "%")
```

A vector is either numeric or character, but never both (see how the 7 was changed to "7").

You can turn one into the other (if possible) as follows:

```{r}
x <- 1:10
x
as.character(x)
```

```{r}
x <- c("1", "5", "10", "-3")
x
as.numeric(x)
```

A third type of data is logical, with values either TRUE or FALSE. 

```{r}
x <- 1:10
x
x > 4
```

these are often used as conditions:

```{r}
x[x>4]
```

This, as we will see shortly, is EXTREMELY useful!

### Data Frames

data frames are the basic format for data in R. They are essentially vectors of equal length put together as columns. 

A data frame can be created as follows:

```{r}
df <- data.frame(
  Gender=c("M", "M", "F", "F", "F"),
  Age=c(23, 25, 19, 22, 21),
  GPA=c(3.5, 3.7, 2.9, 2.8, 3.1)
)
df
```

### Lists

The most general data structures are lists. They are simply a collection of objects. There are no restrictions on what those objects are.

#### **Example**

```{r}
lst <- list(
  Gender=c("M", "M", "F", "F", "F"),
  Age=c(23, 25, 19, 22, 21, 26, 34),
  f=function(x) x^2,
  list(A=c(1, 1), B=c("X", "X", "Y"))
)
lst
```

A data frame is a list with an additional requirement, namely that the elements of the list be of equal length.

#### **Case Study**: UPR Admissions

consider the  **upr** data set . This is the application data for all the students who applied and were accepted to UPR-Mayaguez between 2003 and 2013. 

```{r}
dim(upr)
```

tells us that there were 23666 applications and that for each student there are 16 pieces of information.

```{r}
colnames(upr)
```

shows us the variables 

``` {r}
head(upr, 3)
```

shows us the first three cases.

Let's say we want to find the number of males and females. We can use the table command for that:

```{r,echo=FALSE}
rm(Gender)
```


``` {r, eval=FALSE}
table(Gender)
```

```{r, echo=FALSE}
cat("Error: object 'Gender' not found\n")
```


What happened? Right now R does not know what Gender is because it is "hidden" inside the upr data set. Think of **upr** as a box that is currently closed, so R can't look inside and see the column names. We need to open the box first:

```{r}
attach(upr)
table(Gender)
```

there is also a detach command to undo an attach, but this is not usually needed because the attach goes away when you close R.  

**Note**: you need to attach a data frame only once in each session working with R. 

**Note**: Say you are working first with a data set "students 2016" which has a column called Gender, and you attached it. Later (but in the same R session) you start working with a data set "students 2017" which also has a column called Gender, and you are attaching this one as well. If you use Gender now it will be from "students 2017".

### Subsetting of Data Frames

Consider the following data frame (not a real data set):

```{r, echo=FALSE}
set.seed(111)
Age <- sample(18:25, size=10, replace=T)
GPA <- round(runif(10, 2, 4), 1)
Gender <- sample(c("Male", "Female"), size=10, replace=T)
students <- data.frame(Age, GPA, Gender)
```

```{r}
students
```

Here each single piece of data is identified by its row number and its column number. So for example in row 2, column 2 we have "3.2", in row 6, column 3 we have "Male". 

As with the vectors before we can use the  [ ] notation to access pieces of a data frame, but now we need to give it both the row and the column number, separated by a ,:

```{r}
students[6, 3]
```

As before we can pick more than one piece:

```{r}
students[1:5, 3]
students[1:5, 1:2]
students[-c(1:5), 3]
students[1, ]
students[, 2]
students[, -3]
```

another way of subsetting a data frame is by using the \$ notations:

```{r}
students$Gender
```


### Subsetting of Lists

The double bracket and the \$ notation also work for lists:

#### **Example**

```{r}
lst <- list(
  Gender=c("M", "M", "F", "F", "F"),
  Age=c(23, 25, 19, 22, 21, 26, 34),
  f=function(x) x^2,
  list(A=c(1, 1), B=c("X", "X", "Y"))
)
lst[[4]][[2]]
lst$Gender
```


### Vector Arithmetic 

R allows us to apply any mathematical functions to a whole vector:

```{r}
x <- 1:10
2*x
```

```{r}
x^2
```

```{r}
log(x)
```

```{r}
sum(x)
```

```{r}
y <- 21:30
```

```{r}
x+y
```

```{r}
x^2+y^2   
```

```{r}
mean(x+y) 
```
Let's try something strange:

```{r}
c(1, 2, 3) + c(1, 2, 3, 4)
```
so R notices that we are trying to add a vector of length 3 to a vector of length 4. This should not work, but it actually does!

When it runs out of values in the first vector, R simply starts all over again. 

In general this is more likely a mistake by you, check that this is what you really wanted to do!


### *apply*

A very useful routine in R is *apply*, and its brothers. 

Let's say we have the following matrix:

```{r}
Age <- matrix(sample(20:30, size=100, replace=TRUE), 10, 10)
Age[1:5, 1:5]
```

and we want to find the sums of the ages in each column. Easy:

```{r}
sum(Age[, 1])
sum(Age[, 2])
```
...
```{r}
sum(Age[, 10])
```

or much easier

```{r}
apply(Age, 2, sum)
```

There are a number of apply routines for different data formats. 

#### **Case Study**: upr admissions

Let's say we want to find the mean Highschool GPA:

```{r}
mean(Highschool.GPA)
```

But what if we want to do this for each year separately? Notice that *apply* doesn't work here because the Years are not in separated columns. Instead we can use

```{r}
tapply(Highschool.GPA, Year, mean)
```

